---
title: "Matrix Algebra in R - A Minimal Introduction"
author: "Robert A. Stevens"
date: "November 24, 2014"
output: html_document
---

by James H. Steiger  
Department of Psychology and Human Development  
Vanderbilt University  
Multilevel Regression Modeling, 2009

## Outline

1. Defining a Matrix in R

- Entering by Columns  
- Entering by Rows  
- Entering a Column or Row Vector

2. Extracting Pieces of a Matrix

- Extracting Individual Elements  
- Extracting a Row of a Matrix  
- Extracting a Column of a Matrix  
- Extracting Several Rows and/or Columns

3. Combining Matrices 

- Joining Rows  
- Joining Columns

4. Basic Matrix Operations

- Matrix Addition and Subtraction Scalar Multiplication  
- Matrix Multiplication  
- Matrix Transposition  
- Matrix Inversion

## Preliminary Comments

- This is a very basic introduction

- For some more challenging basics, you might examine Chapter 5 of *An Introduction to R*,  the manual available from the Help->PDF Manuals menu selection in the R program

## Entering a Matrix

- Suppose you wish to enter, then view the following matrix A in R

> 1 2  
> 3 4

- You would use the R commands:

```{r comment=NA}
A <- matrix(c(1, 3, 2, 4), 2, 2)
A
```

- Note that the numbers are, by default, entered into the matrix *columnwise*, i.e., by column

## Entering a Matrix By Rows

- You can enter the numbers by row, simply by adding an optional input variable

- Here are the R commands:

```{r comment=NA}
A <- matrix(c(1, 2, 3, 4), 2, 2, byrow = TRUE)
A
```

## Entering a Column Vector

- To enter a p × 1 column vector, simply enter a p × 1 matrix 

```{r comment=NA}
a <- matrix(c(1, 2, 3, 4), 4, 1)
a
```

- Row vectors are, likewise, entered as 1 × q matrices

## Extracting Individual Elements

- Individual elements of a matrix are referred to by their subscripts

- For example, consider a matrix correlation matrix R given below

- To extract element R3,1, we simply request R[3, 1]

```{r comment=NA}
r <- c(1.00, 0.40, 0.30, 0.30, 0.40, 1.00, 0.20, 0.20, 0.30, 0.20, 1.00, 0.30, 0.30, 0.20, 0.30, 1.00)
R <- matrix(r, 4, 4)
R
R[3, 1] 
```

## Extracting a Row of a Matrix

- To get an entire row of a matrix, you name the row and leave out the column

- For example, in the matrix R below, to get the first row, just enter R[1, ]

```{r comment=NA}
R[1, ]
```

## Extracting a Column of a Matrix

- To get an entire column of a matrix, you name the column and leave out the row

- For example, in the matrix R below, to get the first column, just enter R[ , 1]

```{r comment=NA}
R[ , 1]
```

## Extracting Several Rows and/or Columns

Examine the following examples to see how we can extract any specified range of rows and/or columns

```{r comment=NA}
R[1:3, ]
R[1:3, 2:4]
```

## Joining Rows

- On occasion, we need to build up matrices from smaller parts

- You can combine several matrices with the same number of columns by joining them as rows, using the rbind() command

- Here is an example

```{r comment=NA}
A <- matrix(c(1, 3, 3, 9, 6, 5), 2, 3) 
B <- matrix(c(9, 8, 8, 2, 9, 0), 2, 3)
A
B
rbind(A, B)
rbind(B, A)
```

## Joining Columns

- In similar fashion, you can combine several matrices with the same number of rows by joining them as columns, using the cbind() command

- Here is an example

```{r comment=NA}
A <- matrix(c(1, 3, 3, 9, 6, 5), 2, 3) 
B <- matrix(c(9, 8, 8, 2, 9, 0), 2, 3)
A
B
cbind(A, B)
cbind(B, A)
```

## Matrix Addition and Subtraction

Adding or subtracting matrices is natural and straightfoward, as the example below shows

```{r comment=NA}
A <- matrix(c(1, 3, 3, 9), 2, 2) 
B <- matrix(c(9, 8, 8, 2), 2, 2)
A
B
A + B
A - B
```

## Scalar Multiplication

To multiply a matrix by a scalar, simply use the multiplication symbol *

For example,

```{r comment=NA}
A
3*A
```

## Matrix Multiplication

Matrix multiplication uses the %*% command

Example

```{r comment=NA}
A <- matrix(c(1, 3, 3, 9), 2, 2) 
B <- matrix(c(9, 8, 8, 2), 2, 2)
A
B
A %*% B
B %*% A
```

## Matrix Transposition

To transpose a matrix, use the t() command

Example

```{r comment=NA}
A <- matrix(c(1, 3, 3, 9, 6, 5), 2, 3) 
B <- matrix(c(9, 8, 8, 2, 9, 0), 2, 3)
A
B
t(A)
t(B)
```

## Matrix Inversion

- To invert a square matrix, use the solve() command

- In the example below, we illustrate a common problem — numbers that are really zero are only very close to zero due to rounding error

- When we compute the product AA−1, we should get the identity matrix I , but instead we see that the off-diagonal elements are not quite zero.

- To cure this problem, you can use the zapsmall() function

Example 

```{r comment=NA}
A <- matrix(c(1, 9, 9, 3, 6, 1, 3, 5, 8), 3, 3, byrow = TRUE) 
A
solve(A)
A %*% solve(A)
zapsmall(A %*% solve(A))
```
